How to improve your relationship with your future self.

### 2016-06-24 workflow.Rmd
more changes, time for dinner!

Moreover, coding saves time. Often, a lot of time, especially if projects increase (in the number of files, partners, analysis). Manually importing
a data file into R, may be effortless. Importing 100 files is another issue and the costs of each manual action add up quickly.

@@

NOTE we could include some code here to provide an example of automation of results?

@@

Data analysis involves a long series of decisions. Each decision
requires a justification.  Some decisions will be too small and
technical for inclusion in the published article itself. These need to
be documented in the code itself [@nagler1995coding]. Paragraphs
and citations in the publication will justify the most important
decisions. So, one must code to communicate with yourself and
others. There are two main ways to avoid forgetting the reasons you
did something with data: comment your code and tightly link your code
with your writing.^[One can also try the R command
  `put-it-in-my-mind(reason,importance="high")` to firmly place a
  reason for a decision into the mind of the analyst. I myself have
  not had much luck with this function.]

@@

Comments, unexecuted text inside of a script, are a message to
collaborators (including your future self) and other consumers of your
work. In the above code chunk, I used comments to explain the lines to
readers unfamiliar with R and to remember that I had tried a different
specification but decided not to use it because adding the squared
term did not really change the substantive story arising from the
model.^[R considers text marked with `#` as a comment.]
Messages left for your future self (or near-future others) help
retrace and justify your decisions as the work moves from seminar
paper to conference paper to poster back to paper to dissertation and
onwards.

@@

More importantly, we and others want to know why we did what we
did. Such explanations may not be very clear if we have some pages of
printed code in one hand and a manuscript in the other. Keep in mind
the distinction between the 'source code' of a document (i.e. what
computation was required to produce it) and the visible, type-set page
image. Page images are great for reading, but not great for
reproducing or collaborating. The source code of any document
exchanged by the group must be available and executable.^[As of the 2016 version of this paper, this idea is now widespread and made much easier than before via online services for code sharing and collaboration such as [GitHub](http://github.com) and [BitBucket](http://bitbucket.org).]

@@

For example, in [@sec:data-analys-comp], I suggested that we know
where 'that plot on page 10 'comes from by making sure we had a
`fig1.pdf` file produced from a clearly commented plain text file
called something like `fig1.R`.  An even easier solution would be to
directly include a chunk of code to produce the figure inside of the
paper itself. This paper, for example, was written in plain text using
markdown markup with R code chunks to make things like
Figure \ref{fig:giniprot}. This combination of markdown and R is called
[R Markdown](http://rmarkdown.rstudio.org).

@@

Literate programming need not go against the principle of modular data
analysis [@nagler1995coding]. In my own work I routinely have
several different Sweave files that fulfill different functions, some
of them create \LaTeX code that I can `\input` into my
`main.tex` file, others setup the data, run simulations, or allow
me to record my journeys down blind alleys. Of course, when we have
flying cars running on autopilot, perhaps something other than
Sweave will make our lives even easier. Then we'll change.

**Step 2**
We analyze data in order to explain something about the world to other
scholars and policy makers. If we focus on explaining how we got our
computers to do data analysis to human beings, we will do a better job
with the data analysis itself: we will learn as we focus on teaching,
and we will avoid errors and save time as we ensure that others
(including our future selves) can retrace our steps. A document than
can be 'run' to reproduce all of the analyses also instills
confidence in readers and can more effectively spur discussion and
learning and cumulation of research.

@@

All files containing commands operating on data must refer to a data
file. A reference to a data file is a line of code the analysis
program will use to operate on ('load'/ 'open '/ 'get '/
'use') the data file. One should not have to edit this line on
different computers or platforms in order to execute this command.
Using R, for example, all analysis files should have
`load("thedata.rda")`or `read.csv("thedata.csv")`or some
equivalent line in them, and `thedata.csv`should be stored in
some place easy to find (like in the same directory as the file or
perhaps in `"Data/thedata.rda"`). Of course, it never hurts to
drop in a comment pointing to the data file.

@@

The fact that I need to articulate this idea at all arises because of
graphical user interfaces: it is all too easy to use the mouse to load
a data file into memory and then to write a script to analyze this
file without ever noting the actual name or location of the data file.

@@

Anyone writing custom code should worry about getting it right. The more code one writes, the more time one has to appreciate problems arising from bugs, errors, and typos in data analysis
and code.

Now, if one has a moment to think in between teaching that new class,
reading books for an awards committee, evaluating application files
for the admissions committee, feeding popsicles to a sick child, and
undertaking the odd bit of research, one might say to oneself,
"Before I write new code, I should write a test of the code. I should
write a little bit of code that lets me know that my double-bootstrap
procedure actually does what it is supposed to do."

Of course, this idea, like most others, is not new. The desire to
avoid error looms large when large groups of programmers write code
for multi-million dollar programs. The idea of
[test driven development](http://en.wikipedia.org/wiki/Test-driven_development) and the idea that one ought to create tests of
[small parts of one's code](http://en.wikipedia.org/wiki/Unit_testing) arose to address such concerns.^[There are now R packages to help R package writers do this, see for example <https://github.com/hadley/testthat> and the article on it <https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf>.] For the social scientist
collaborating with her future self and/or a small group of
collaborators, here is an example of this idea in a very simple form:
Say I want to write a function to multiply a number by 2. If my
function works, when I give it the number 4, I should see it return
the number 8 and when I give it -4, I should get -8.

@@

Ack! I mistyped '+' for '*'. Good thing I wrote the
test!^[A more common example of this kind of testing occur
  everyday when we recode variables into new forms but look at a
  crosstab of the old vs. new variable before proceeding.]

@@

Lots of people are thinking about 'reproducible research'and
'literate programming' these days. Google those terms. Of course, the
devil is in the details: Here I list a few of my own attempts at enabling
reproducible research. You'll find many other inspiring examples on
the web. Luckily, the open source ethos aligns nicely with academic
incentives, so we are beginning to find more and more people offering
their files online for copying and improvement. By the way, if you do
copy and improve, it is polite to alert the person from whom you made
the copy about your work.

@@

**Step 6** We all learn by doing. When we share
reproduction materials we improve both cumulation of knowledge and our
methods for doing social science
[@freese2007replication,king1995replication}.  As we copy and
improve upon each other's modes of doing work we enhance our collective
ability to believe each other and for future scholars to believe us, too.

@@

But what if no one ever hears of your work, or, by some cruel fate, your
article does not spawn debate? Why then would you spend time to
communicate with your future self and others? My own answer to this
question is that I want my work to be credible and useful to myself
and other scholars even if each article does not immediately change
the world.  What I report in my data analyses should have two main
characteristics: (1) the findings of the work should not be a matter
of opinion; and (2) other people should be able to reproduce the findings. That is,
the work represents a shared
experience --- and an experience shared without respect to the
identities of others (although requiring some common technical
training and research resources).

Assume we want others to believe us when we say something. More narrowly,
assume we want other people to believe us when we say something about data:
'data' here can be words, numbers, musical notes, images, ideas, etc \ldots The
point is that we are making some claims about patterns in some collection of
stuff. For example, when I was invited into the homes and offices of ordinary
people in Chile in 1991, the "stuff" was recordings of conversations that we
had about life during the first year of democracy and I was comparing the
responses of people who had experienced the Pinochet dictatorship differently.
Now, it might be easy to convince others that 'this collection of stuff is
different from that collection of stuff' if those people were looking over our
shoulders the whole time that we made decisions about collecting the stuff and
broke it up into understandable parts and reorganized and summarized it.
Unfortunately, we can't assume that people are willing to shadow a researcher
throughout her career. Rather, we do our work alone or in small groups and want
to convince other distant and future people about our analyses.

>>>>

Moreover, coding saves time. Often, a lot of time, especially if projects increase (in the number of files, partners, analysis). Manually importing
a data file into R, may be effortless. Importing 100 files is another issue, and the costs of each manual action add up quickly.

@@

[NOTE we could include some code here to provide an example of automation of results?]

@@

Data analysis involves a long series of decisions. Each decision
requires a justification.  Some decisions will be too small and
technical for inclusion in the published article itself. These need to
be documented in the code itself [@nagler1995coding]. Paragraphs
and citations in the publication will justify the most important
decisions. So, one must code to communicate with yourself and
others. There are two main ways to avoid forgetting the reasons you
did something with data: comment your code and tightly link your code
with your writing.^[One can also try the R command
  `put-it-in-my-mind(reason,importance="high")` to firmly place a
  reason for a decision into the mind of the analyst. I myself have
  not had much luck with this function.] [I FEEL THIS COMMENT IS FUNNY BUT CONFUSING FOR OUR LATIN AMERICAN READERS PROBABLY]

@@

Comments, unexecuted text inside of a script, are a message to
collaborators (including your future self) and other consumers of your
work. In the above code chunk, we used comments to explain the lines to
readers unfamiliar with R and to remember that we had tried a different
specification but decided not to use it because adding the squared
term did not really change the substantive story arising from the
model.^[R considers text marked with `#` as a comment.]
Messages left for your future self (or near-future others) help
retrace and justify your decisions as the work moves from seminar
paper to conference paper to poster back to paper to dissertation and
onwards.

@@

More importantly, we and others want to know why we did what we
did. Such explanations may not be very clear if we have some pages of
printed code in one hand and a manuscript in the other. Keep in mind
the distinction between the 'source code' of a document (i.e. what
computation was required to produce it) and the visible, type-set page
image. Page images are great for reading, but not great for
reproducing or collaborating. The source code of any document
exchanged by the group must be available and executable.^[As of the 2016 version of this paper, this idea is now widespread and made much easier than before via online services for code sharing and collaboration such as [GitHub](http://github.com) and [BitBucket](http://bitbucket.org).]
[I FEEL WE CAN MOVE THIS TO THE MAIN TEXT AND (BRIEFLY) INTRODUCE GITHUB]

@@

For example, in [@sec:data-analys-comp], I suggested that we know
where 'that plot on page 10 'comes from by making sure we had a
`fig1.pdf` file produced from a clearly commented plain text file
called something like `makefig1.R`.  An even easier solution would be to
directly include a chunk of code to produce the figure inside of the
paper itself. This paper, for example, was written in plain text using
markdown markup with R code chunks to make things like
Figure \ref{fig:giniprot}. This combination of markdown and R is called
[R Markdown](http://rmarkdown.rstudio.org).

@@

Literate programming need not go against the principle of modular data
analysis [@nagler1995coding]. Jake routinely uses
several different Sweave files that fulfill different functions, some
of them create \LaTeX code that he can `\input` into his
`main.tex` file, others setup the data, run simulations, or allow
him to record his journeys down blind alleys. Of course, when we have
flying cars running on autopilot, perhaps something other than
Sweave will make our lives even easier. Then we'll change.

**Step 2**
We analyze data in order to explain something about the world to other
scholars and policy makers. If we focus on explaining how we got our
computers to do data analysis, we will do a better job
with the data analysis itself: we will learn as we focus on teaching,
and we will avoid errors and save time as we ensure that others
(including our future selves) can retrace our steps. A document than
can be 'run' to reproduce all of the analyses also instills
confidence in readers and can more effectively spur discussion and
learning and cumulation of research.

@@

All files containing commands operating on data must refer to a data
file. A reference to a data file is a line of code the analysis
program will use to operate on ('load'/ 'open '/ 'get '/
'use') the data file. One should not have to edit this line on
different computers or platforms in order to execute this command.
Using R, for example, all analysis files should have
`load("thedata.rda")`or `read.csv("thedata.csv")`or some
equivalent line in them, and `thedata.csv`should be stored in
some place easy to find (like in the same directory as the file or
perhaps in `"Data/thedata.rda"`). Of course, its even better to
include a comment pointing to the data file.

@@

The fact that we need to articulate this idea at all arises because of
graphical user interfaces: it is all too easy to use the mouse to load
a data file into memory and then to write a script to analyze this
file without ever noting the actual name or location of the data file.

@@

Anyone writing custom code should worry about getting it right. The more code one writes, the more time one has to appreciate problems arising from bugs, errors, and typos in data analysis and code. Mistakes will always be made we should just recognize we need more standardization of practice to minimise the mistakes and help others find them. One thing we can do is to include testing in our coding.

If one has a moment to think in between teaching that new class,
reading books for an awards committee, evaluating application files
for the admissions committee, feeding popsicles to a sick child, and
undertaking the odd bit of research, one might say to oneself,
"Before I write new code, I should write a test of the code. I should
write a little bit of code that lets me know that my double-bootstrap
procedure actually does what it is supposed to do."

Of course, this idea, like most others, is not new. The desire to
avoid error looms large when large groups of programmers write code
for multi-million dollar programs. The idea of
[test driven development](http://en.wikipedia.org/wiki/Test-driven_development) and the idea that one ought to create tests of
[small parts of one's code](http://en.wikipedia.org/wiki/Unit_testing) arose to address such concerns.^[There are now R packages to help R package writers do this, see for example <https://github.com/hadley/testthat> and the article on it <https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf>.] For the social scientist
collaborating with her future self and/or a small group of
collaborators, here is an example of this idea in a very simple form:
Say you want to write a function to multiply a number by 2. If the
function works, when you give it the number 4, you should see it return
the number 8 and when you give it -4, you should get -8.

@@

Ack! we mistyped '+' for '*'. Good thing I wrote the
test!^[A more common example of this kind of testing occur
  everyday when we recode variables into new forms but look at a
  crosstab of the old vs. new variable before proceeding.]

@@

[IS THIS SECTION MORE MORE ON CREATING AN INTEGRATED REPRODUCABLE WORKFLOW RATHER THAN COPYING AND IMPROVING?]
Lots of people are thinking about 'reproducible research', creating a 'reproducable workflow' and
'literate programming' these days. Google those terms. Of course, the
devil is in the details: Here I list a few of my own attempts at enabling
reproducible research. You'll find many other inspiring examples on
the web. Luckily, the open source ethos aligns nicely with academic
incentives, so we are beginning to find more and more people offering
their files online for copying and improvement. By the way, if you do
copy and improve, it is polite to alert the person from whom you made
the copy about your work.

@@

[ADD OPEN SCIENCE FRAMEWORK, COLLABORATING ON DROPBOX AND NEW INITIATIVES LIKE Dynamic Documents (R Markdown/KnitR in R Studio,
Markdoc in Stata), Jupyter Notebook, Docker for Cross-Platform Reproducibility)]

**Step 6** We all learn by doing. When we create a reproducible workflow and share
reproduction materials we improve both cumulation of knowledge and our
methods for doing social science
[@freese2007replication,king1995replication}.  As we copy and
improve upon each other's modes of doing work we enhance our collective
ability to believe each other and for future scholars to believe us, too.
[I FEEL LIKE THE COPY COMMENT IS PERHAPS ANOTHER POINT TO MAKE (OR LEAVE)]

@@

But what if no one ever hears of your work, or, by some cruel fate, your
article does not spawn debate? Why then would you spend time to
communicate with your future self and others? Our own answer to this
question is that we want our work to be credible and useful to ourself
and other scholars regardless.  What we report in our data analyses should have two main
characteristics: (1) the findings of the work should not be a matter
of opinion; and (2) other people should be able to reproduce the findings. That is,
the work represents a shared experience --- and an experience shared without respect to the
identities of others (although requiring some common technical
training and research resources).

Assume we want others to believe us when we say something. More narrowly,
assume we want other people to believe us when we say something about data:
'data' here can be words, numbers, musical notes, images, ideas, etc \ldots The
point is that we are making some claims about patterns in some collection of
stuff. For example, when Jake was invited into the homes and offices of ordinary
people in Chile in 1991, the "stuff" was recordings of conversations that we
had about life during the first year of democracy and Jake was comparing the
responses of people who had experienced the Pinochet dictatorship differently.
Now, it might be easy to convince others that 'this collection of stuff' is
different from 'that collection of stuff' if those people were looking over our
shoulders the whole time that we made decisions about collecting the stuff and
broke it up into understandable parts and reorganized and summarized it.
Unfortunately, we can't assume that people are willing to shadow a researcher
throughout her career. Rather, we do our work alone or in small groups and want
to convince other distant and future people about our analyses and findings.
